---
title: "Newport Beach Analysis"
author: "Christine Zhang"
date: "3/6/2019"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

An analysis of race/ethnicity categories in Newport Beach, California.

### Load libraries and read in data

The data are from the Census Bureau's 2013-2017 American Community Survey (Five-Year Estimates), table B03002, [HISPANIC OR LATINO ORIGIN BY RACE](https://factfinder.census.gov/bkmk/table/1.0/en/ACS/17_5YR/B03002/1600000US0651182). The numbers represent estimates over the five year period.

```{r loading, warning = FALSE, message = FALSE}
library('tidyverse')
library('knitr')
newport.beach <- read_csv('input/newport_beach.csv')
```

### Create race/ethnicity categories

Rename the race/ethnicity variables we want to keep into more intuitive names; e.g., "white" for non-Hispanic white, "black" for non-Hispanic black, etc. The "other" category includes non-Hispanic American Indian and Alaskan Native, Native Hawaiian and Other Pacific Islander, Some other race alone, and Two or more races.

```{r creating_cats}

# v17 <- load_variables(2017, "acs5", cache = TRUE) # view variable names

newport.beach <- newport.beach %>% select(-summary_moe) %>%
  mutate(cat = case_when(
    variable == 'B03002_003' ~ "white",
    variable == 'B03002_004' ~ "black",
    variable == 'B03002_005' | variable == 'B03002_007' | variable == 'B03002_008' | variable == 'B03002_009' ~ "other",
    variable == 'B03002_006' ~ "asian",
    variable == 'B03002_012' ~ "hisp")) 
```

### Filter out unwanted categories and summarize the data by the newly-created categories

```{r summarizing}
newport.beach <- newport.beach %>% 
  filter(!is.na(cat)) %>% 
  group_by(GEOID, NAME, cat, summary_est) %>% 
  summarise(estimate = sum(estimate))
```

### Calculate each category's percentage of total population and sort categories by percentage, from highest to lowest

We use the `kable()` function from the `knitr` package to render a formatted table.

```{r percentages}
newport.beach <- newport.beach %>% 
  mutate(perc = estimate/summary_est * 100) %>% 
  arrange(desc(perc))

kable(newport.beach) # formatted table
```

### Write to a CSV

```{r write}
write_csv(newport.beach, 'output/newport_beach_cleaned.csv')
```